---
title: 'What car characteristics influence Fuel Efficiency? A Data Visualization'
author: "Laura Brune"
date: "2026-04-22"
format: 
  html: 
    toc: true 
    number-sections: true
    css: styles.css
    theme: "flatly"
    toc-location: left
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```


```{r}
#| label: load-main


library(here)
source(here("R", "main.R"))
```


This report investigates with data visualizations which characteristics influence fuel efficiency of cars on the American market between 1984 and 2014. The variables analysed are: type of gear, number of gears, number of cylinders and brand. The data is from...


# Fuel Efficiency over Time 

```{r}
#| label: "fig-fuel-efficiency-over-time"
#| fig-cap: "Fuel efficiency over Time"


f_time
```

-   in 1985 the average fuel consumption of an American car was 14.3 liters in the city and 13.3 liters on the highway 
-   in 2014 the average fuel consumption was 12.4 liters in the city and 8.9 liters on the highway
-   although both measurements decreased overtime, fuel efficiency increased higher on the highway
-   between 1985 and 2006 fuel consumption stayed on a similar level
-   only since 2006 cars on the American market got more efficient 



# Automatic Gear

```{r}
#| label: "fig-automatic-gear-per-year"
#| fig-cap: "Relationship between the Share of Automatic Gear and Fuel Efficiency per Year"
#| fig-height: 10


f_automatic_year_total
```

-   automatic gearboxes were a popular development in the American car market 
-   by the mid-1980s, more than half of American cars had automatic gears 
-   by 2014, the share of automatic gearboxes increased to more than four fitht 
-   Figure 3 shows a relationship between a higher share of automatic gearboxes and fuel efficiency: the higher the share of automatic gearboxes the higher is the average fuel efficiency
-   however this relationship is non-linear and intensifies in the late 2000s

# Number of Gears

```{r}
#| label: "fig-n-gears"
#| fig-cap: "Relationship between the Number of Gears and Fuel Efficiency"
#| fig-height: 8


f_nGears_total
```

```{r}
#| label: "fig-n-gears-year"
#| fig-cap: "Mean and Predominant Number of Gears, \n Time and Fuel Efficiency"
#| fig-height: 8


f_nGears_annual_total
```

-   more gears are associated with better fuel efficiency
-   since 2009 six gears predominate the American market


# Number of Cylinders

```{r}
#| label: "fig-n-cylinders"
#| fig-cap: "Relationship between the Number of Cylinders \n and Fuel Efficiency"
#| fig-height: 8


f_Cylinders
```

-   with each additional cylinder fuel efficiency decreases 
-   except for 12 cylinders in the city and 12 and 10 cylinders on the highway


# Brand 
```{r}
#| label: "fig-top-12-brand"
#| fig-cap: "Relationship between Fuel Efficiency in the City \n and on the Highway for the Top 12 Brands"


f_make
```

-  brands differ highly in the distribution of their models
-  Volvo and Subarus fuel efficiency is very similar between models 
-  brands like Dodge and Chevrolet have models with a wide range of fuel consumptions  

# Conclusion

**Key Findings**

-  fuel efficiency increased since the late 2000s
-  a higher share of automatic gears is associated with better fuel efficiency 
-  cars with a higher number of gears have in general a better fuel efficiency 
-  a lower number of cylinders is associated with a higher fuel efficiency 

**Limitations**

-  only descriptive analysis is used 
-  to confirm results a technical analysis is needed 

# Data Source 

cite data 
